Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-01-26 paper/aldebaran.tex
Minor edits to wording, some references back and forth to Appendix 1.

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process, with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. Gaussian processes with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, this is impractical for long time series, as the computational
cost of evaluating the standard Gaussian Process likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the solution of a class of stochastic ordinary differential equations, the Continuous Auto-Regressive Moving Average
(CARMA) models, whose likelihoods can be evaluated in linear time. For problems
such as this, we therefore have access to these
powerful computational tools for inferring their power spectral densities.
\citet{Kelly2014} gives more details on CARMA processes in an astronomical
context, and shows how a state-space model of the solution $y(t)$ can be tracked
through a time-series of uncertain observations $y_k = y\left( t_k \right) +
\epsilon_k$, with Gaussian noise $\epsilon_k$ of zero mean and known
(heteroskedastic) variance, using a Kalman filter to produce a likelihood
function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on the amplitude
of the stochastic forcing, $\sigma$, the (possibly complex) eigenfrequencies of
the ODE, $r_i$, and parameters describing the power spectrum of the stochastic
forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational time.
Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion can be
used to reduce the standard Gaussian Process covariance matrix for this same
process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl}
package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman
filters that can compute the likelihood for a set of observations parameterised
by either the $r_i$ and $b_j$ or the RMS amplitudes $A_i$ of each eigenmode of
the ODE and the corresponding roots $r_i$.  We also include a Keplerian RV
signal from a planetary companion to Aldebaran \citep{Hatzes2015,Gregory2005}.
Because we have a set of observations taken with different instruments at
different sites (see Section \ref{sec:archival-obs}), we include RV offset
(i.e.\ mean value) parameters that are instrument- and site-dependent and also
an instrument- and site-dependent scaling parameter.  Thus, the $i$th RV
measured at site and intstrument combination $j$, $y_{ij}$, is assumed to be
%
\begin{equation}
  \label{eq:RV-model}
  y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j
\epsilon_{ij},
\end{equation}
%
where $t_i$ is the time of the measurement, $v_r$ is the planetary RV signal,
$\mu_j$ is the mean RV offset for site and instrument combination $j$, $y(t)$ is
the CARMA process representing stellar activity, $\nu_j$ is the uncertainty
scaling parameter for site and instrument combination $j$, and $\epsilon_{ij}$
is an independent Gaussian random variable with mean zero and standard deviation
equal to the quoted uncertainty of the observation.  We impose broad priors (see
Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the
CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.  We treat the
frequency of maximum asteroseismic power, $\nu_\mathrm{max}$, as the frequency
of an appropriate eigenmode in the CARMA process representing the stellar
activity; this is equivalent to fitting a Lorentzian profile to a set of
unresolved asteroseismic modes.  We used the nested sampling algorithm of the
\texttt{Ensemble.jl} package\footnote{\url{https://github.com/farr/Ensemble.jl};
this package implements several stochastic sampling algorithms based on the
``stretch move'' proposal used in \texttt{emcee} \citep{Foreman-Mackey2013} and
introduced in \citet{Goodman2010}} to calculate the marginal likelihood
(evidences) and draw samples from the posterior distribution over the parameters
of our various models.  The priors we impose on the parameters of our model are given in Table \ref{tab:parameters}.

@@

\subsection{Planetary and Stellar Parameters}

@@

The priors we impose on the parameters of the CARMA model, planetary RV signal,
and instrumental and site effects are given in Table \ref{tab:parameters}.

>>>>

If we apply nonlinear statistical inference methods this situation can be
improved. \citet{brewer2009} show that a system of driven, damped harmonic
oscillators such as we encounter in asteroseismology can be statistically
modelled as a Gaussian Process, with a covariance kernel consisting of a sum of
damped sinusoids. The hyperparameters of this process encode features of the
power spectral density distributions and are insensitive to the window function. Gaussian processes with quasi-periodic kernels have previously been used to re-analyse RV data for main sequence stars \citep{2014MNRAS.443.2517H,2015MNRAS.452.2269R}, where the noise is from stellar activity, but have not previously been applied to red giant asteroseismology.
Unfortunately, these methods are impractical for long time series, as the computational
cost of evaluating the standard Gaussian Process likelihood function scales as
$\mathcal{O}(n^3)$.

Fortunately, the damped and driven harmonic oscillator GP can be written as the
solution of a class of stochastic ordinary differential equations, the
Continuous Auto-Regressive Moving Average (CARMA) models, whose likelihoods can
be evaluated in linear time. For problems such as this, we therefore have access
to these powerful computational tools for inferring their power spectral
densities.  Our treatment of CARMA models is described more fully in Appendix
\ref{carmatheory}.   \citet{Kelly2014} also gives more details on CARMA
processes in an astronomical context, and shows how a state-space model of the
process $y(t)$ can be tracked through a time-series of uncertain observations
$y_k = y\left( t_k \right) + \epsilon_k$, with Gaussian noise $\epsilon_k$ of
zero mean and known (heteroskedastic) variance, using a Kalman filter to produce
a likelihood function $p\left( y_k \mid \sigma, r_i, b_j \right)$ depending on
the amplitude of the stochastic forcing, $\sigma$, the (possibly complex)
eigenfrequencies of the ODE, $r_i$, and parameters describing the power spectrum
of the stochastic forcing, $b_j$ in $\mathcal{O}\left( n \right)$ computational
time.  Additionally, \citet{Foreman-Mackey2017} show how a dimensional expansion
can be used to reduce the standard Gaussian Process covariance matrix for this
same process,  with autocorrelation given by Eq.\ \eqref{eq:CARMA-ACF} under
observations at times $t_k$, to a banded form that can then be
Cholesky-decomposed in the standard Gaussian Process likelihood function in
$\mathcal{O}\left( n \right)$ time.

Here we have adopted the \citet{Kelly2014} approach.  The \texttt{CARMA.jl}
package\footnote{\url{https://github.com/farr/CARMA.jl}} implements Kalman
filters that can compute the likelihood for a set of observations parameterised
by either the $r_i$ and $b_j$ or the RMS amplitudes $A_i$ of each eigenmode of
the ODE and the corresponding roots $r_i$.  We also include a Keplerian RV
signal from a planetary companion to Aldebaran \citep{Hatzes2015,Gregory2005}.
Because we have a set of observations taken with different instruments at
different sites (see Section \ref{sec:archival-obs}), we include RV offset
(i.e.\ mean value) parameters that are instrument- and site-dependent and also
an instrument- and site-dependent scaling parameter.  Thus, the $i$th RV
measured at site and intstrument combination $j$, $y_{ij}$, is assumed to be
%
\begin{equation}
  \label{eq:RV-model}
  y_{ij} = v_r\left(t_i\right) + \mu_j + y\left( t_i \right) + \nu_j
\epsilon_{ij},
\end{equation}
%
where $t_i$ is the time of the measurement, $v_r$ is the planetary RV signal,
$\mu_j$ is the mean RV offset for site and instrument combination $j$, $y(t)$ is
the CARMA process representing stellar activity, $\nu_j$ is the uncertainty
scaling parameter for site and instrument combination $j$, and $\epsilon_{ij}$
is an independent Gaussian random variable with mean zero and standard deviation
equal to the quoted uncertainty of the observation.  We impose broad priors (see
Table \ref{tab:parameters}) on the parameters of the planetary RV signal, the
CARMA process, the $\mu_j$ offsets, and the $\nu_j$ scale factors.  We treat the
frequency of maximum asteroseismic power, $\nu_\mathrm{max}$, as the frequency
of an appropriate eigenmode in the CARMA process representing the stellar
activity; this is equivalent to fitting a Lorentzian profile to a set of
unresolved asteroseismic modes.  We used the nested sampling algorithm of the
\texttt{Ensemble.jl} package\footnote{\url{https://github.com/farr/Ensemble.jl};
this package implements several stochastic sampling algorithms based on the
``stretch move'' proposal used in \texttt{emcee} \citep{Foreman-Mackey2013} and
introduced in \citet{Goodman2010}} to calculate the marginal likelihood
(evidences) and draw samples from the posterior distribution over the parameters
of our various models.

@@

\subsection{Planetary and Stellar Parameters}
\label{sec:planet-star-parameters}

@@

Our preferred model for the data uses one decaying mode and one oscillatory mode
in the CARMA process to capture the un-resolved superposition of many
asteroseismic modes (see Section \ref{sec:planet-star-parameters}); we have fit
models with more than one mode of each type, but find no improvement in
model-selection criteria from the expanded parameter space.  The priors we
impose in all cases on the parameters of the CARMA model, planetary RV signal,
and instrumental and site effects are given in Table \ref{tab:parameters}.
